Normalization by Evaluation for Sized Dependent Types

### 2017-02-27 nbe/icfp17.tex
[ ICFP 17 ] Proof-read the introduction.

By \emph{large eliminations} \cite{werner:types92} we refer to types computed by case distinction over values; they occur
in type theories that feature both universes and data types.
In the presence of large eliminations, dependency erasure fails, and it is not clear what the skeleton of a dependent type should be.  For instance consider the type $\funT n \NN \underbrace{A \to \dots \to A}_n \to A$ of $n$-ary functions; its shape is dependent on the value of $n$, thus cannot be determined statically.
In the solution we present in this article, we do not introduce a new syntax for shapes, just a relation $T \shape S$ stating that type $S$ qualifies as shape for type $T$.  We use a normalization function inspired by normalization-by-evaluation (NbE) that computes the $\eta$-long normal form of the terms under comparison, which we then check for identity.  NbE \cite{bergerSchwichtenberg:lics91,danvy:tdpe} has proven a robust method to decide equality in powerful type theories with non-trivial $\eta$-laws.  It scales to universes and large eliminations \cite{abelCoquandDybjer:lics07}, topped with singleton types or proof irrelevance \cite{abelCoquandPagano:lmcs11}, and even impredicativity \cite{abel:flops10}.  At its heart there are reflection $\up T$ and reification $\down T$ functions directed by type $T$ and orchestrating just-in-time $\eta$-expansion.  Reflection $\up T x$ maps variables $x$ into the realm of values of type $T$ and let's us compute with open terms.  Reification $\down T a$ takes a value $a$ of type $T$ and computes its long normal form.  For instance, the normal form of a closed function $f : U \to T$ would be $\lambda x.\,\down{T} (f\,(\up U x))$, and for its dependently-typed variant $f : \funT x U T[x]$ it would be $\lambda x.\,\down{T[\up U x]} (f\,(\up U x))$.

The central technical observation is that reflection and reification do not need the precise type $T$, they work the same for any shape $S$ of $T$.  For example, when reflecting a variable $x$ into the polymorphic function type $\forall i.\, F\,i$ where $F\,i = \Nat\,i \to \Nat\,i$, we obtain
$(\up{\forall i.\/ F\,i}x)\,i = \up{F\,i} (x\,i)$ for size $i$ and
$(\up{\forall i.\/ F\,i}x)\,j = \up{F\,j} (x\,j)$ for size $j$.  The new types $F\,i$ and $F\,j$ we reflect at are no longer equal (and they are not subtypes of each other), but they still have the same shape, $\Nat\_ \to \Nat\_$.  This means they will still move in lock-step what $\eta$-expansion concerns, which is sufficient to prove NbE correct for judgmental equality.
We call the enabling property of $F$ \emph{shape irrelevance}, meaning that for any pair $i$, $j$ of legal arguments, $F\,i$ and $F\,j$ have the same shape.  Whenever we form a irrelevant function type $\forall x \of U.\, T[x]$, we require $T[x]$ to be shape-irrelevant in $x$.  This is the middle ground between ICC$^*$, where no restriction is placed on $T$ but $\eta$ for unit types is (at least, currently) out of reach, and Pfenning's \citeyearpar{pfenning:intextirr} irrelevance modality, adapted to full dependent types by \citet{abelScherer:types10}, which requires $T$ to be irrelevant in $x$ and, thus, has type equality $T[i] = T[j]$.

For the time being, we do not (and cannot) develop a general theory of shape irrelevance.  We confine ourselves to size-irrelevant function types $\forall i.\, T[i]$.  This relieves us from defining a special shape-irrelevance modality, since all size-indexed types $T[i]$ are shape irrelevant in $i$, simply because there is no case distinction on size, and sizes appear relevantly only under a sized type constructor such as $\tNat$.  Our technique would not extend to the polymorphic types $\forall X.\,B[X]$ of System~F.  Even though there is no case distinction on types, shape irrelevance of $B[X]$ fails in general, as $X$ could appear as a type on the top-level, e.g. in $B[X] = X$ or $B[X] = X \to X$.

This article makes the following novel contributions:
\begin{enumerate}

@@

\item We adapt normalization-by-evaluation (NbE) to sized types and size quantification and show that it decides judgmental equality (sections~\ref{sec:sem} and \ref{sec:soundnbe}).  The novel technical tool is a relation $T \shape S$ which relates a type $T$ to its possible shapes $S$.  This approximation relation allows reflection and reification at size-polymorphic types $\forall i.\, T$.
As usual for the meta-theory of Type Theory with large eliminations, the machinery is involved, but just require the usual two logical relations:  First, a PER model to define the semantics of types and prove the completeness of NbE (Section~\ref{sec:sem}).  Secondly, a relation between syntax and semantics to prove soundness of NbE (Section~\ref{sec:soundnbe}).

\item We present an bidirectional type checking algorithm \cite{coquand:type} which takes the irrelevant size argument as reliable hint for the type checker (sections~\ref{sec:asub} and \ref{sec:tycheck}).  It is complete for normal forms which can be typed with the restricted rule for $\forall$-elimination:
\[
\ru{\Gamma \der t : \forall i.\, T
  }{\Gamma \der t\,a : T[a/i]}
\]
The algorithm employs the usual lazy reduction for types (just-in-time weak-head evaluation) in type and sub-type checker
\cite{huet:wsscs89}.  In this, it improves on \citet{fridlenderPagano:tlca13} which instruments full normalization (NbE) at every step.
\end{enumerate}
This article is accompanied by a prototypical type checker \texttt{Sit} which implements the type system and type checking algorithm as described in the remainder of the paper.
But first, we will motivate our type system from a practical perspective: reasoning about programs involving sized types in Agda.

>>>>

By \emph{large eliminations} \cite{werner:types92} we refer to types computed by case distinction over values; they occur
in type theories that feature both universes and data types.
In the presence of large eliminations, dependency erasure fails, and it is not clear what the skeleton of a dependent type should be.  For instance consider the type $\funT n \NN \underbrace{A \to \dots \to A}_n \to A$ of $n$-ary functions; its shape is dependent on the value of $n$, thus cannot be determined statically.  Thus, the ``skeleton'' idea is also not directly applicable.

Going beyond the classical incremental equality check, there is a technique that can deal with dynamic $\eta$-expansion.  It is a type-directed normalization function inspired by normalization-by-evaluation (NbE) that computes $\eta$-long normal forms \cite{bergerSchwichtenberg:lics91,danvy:tdpe}.  We can check the computed normal forms for identity and, thus, decide definitional equality.
NbE has proven to be a robust method to decide equality in powerful type theories with non-trivial $\eta$-laws.  It scales to universes and large eliminations \cite{abelCoquandDybjer:lics07}, topped with singleton types or proof irrelevance \cite{abelCoquandPagano:lmcs11}, and even impredicativity \cite{abel:flops10}.  At its heart there are reflection $\up T$ and reification $\down T$ functions directed by type $T$ and orchestrating just-in-time $\eta$-expansion.  Reflection $\up T x$ maps variables $x$ into the realm of values of type $T$ and let's us compute with open terms.  Reification $\down T a$ takes a value $a$ of type $T$ and computes its long normal form.  For instance, the normal form of a closed function $f : U \to T$ would be $\lambda x.\,\down{T} (f\,(\up U x))$, and for its dependently-typed variant $f : \funT x U T[x]$ it would be $\lambda x.\,\down{T[\up U x]} (f\,(\up U x))$.

% In the solution we present in this article, we do not introduce a new syntax for shapes, just a relation $T \shape S$ stating that type $S$ qualifies as shape for type $T$.  We use a normalization function inspired by normalization-by-evaluation (NbE) that computes the $\eta$-long normal form of the terms under comparison, which we then check for identity.

The central technical observation is that reflection and reification do not need the precise type $T$, they work the same for any \emph{shape} $S$ of $T$.
We managed, while not introducing a new syntax for shapes, to define a relation $T \shape S$ on type values stating that type $S$ qualifies as shape for type $T$.
Shapes unfold dynamically during reflection and reification.
For example, when reflecting a variable $x$ into the polymorphic function type $\forall i.\, F\,i$ where $F\,i = \Nat\,i \to \Nat\,i$, we obtain
$(\up{\forall i.\/ F\,i}x)\,i = \up{F\,i} (x\,i)$ for size $i$ and
$(\up{\forall i.\/ F\,i}x)\,j = \up{F\,j} (x\,j)$ for size $j$.  The new types $F\,i$ and $F\,j$ we reflect at are no longer equal (and they are not subtypes of each other), but they still have the same shape, $\Nat\_ \to \Nat\_$.  This means they will still move in lock-step what $\eta$-expansion concerns, which is sufficient to prove NbE correct for judgmental equality.
We call the enabling property of $F$ \emph{shape irrelevance}, meaning that for any pair $i$, $j$ of legal arguments, $F\,i$ and $F\,j$ have the same shape.  Whenever we form a irrelevant function type $\forall x \of U.\, T[x]$, we require $T[x]$ to be shape-irrelevant in $x$.  This is the middle ground between ICC$^*$, where no restriction is placed on $T$ but $\eta$ for unit types is out of reach (at least for the moment), and Pfenning's \citeyearpar{pfenning:intextirr} irrelevance modality, adapted to full dependent types by \citet{abelScherer:types10}, which requires $T$ to be irrelevant in $x$ and, thus, has type equality $T[i] = T[j]$.

For the time being, we do not (and cannot) develop a general theory of shape irrelevance.  We confine ourselves to size-irrelevant function types $\forall i.\, T[i]$.  This relieves us from defining a special shape-irrelevance modality, since all size-indexed types $T[i]$ \emph{are} shape irrelevant in $i$, simply because there is no case distinction on size, and sizes appear relevantly only under a sized type constructor such as $\tNat$.  Our technique would not extend to the polymorphic types $\forall X.\,B[X]$ of System~F.  Even though there is no case distinction on types, shape irrelevance of $B[X]$ fails in general, as $X$ could appear as a type on the top-level, e.g. in %$B[X] = X$ or
$B[X] = X \to X$, and then $B[\One]$ and $B[A \to A]$ would have distinct shapes.

To summarize, this article makes the following novel contributions:
\begin{enumerate}

@@

\item We adapt normalization-by-evaluation (NbE) to sized types and size quantification and show that it decides judgmental equality (sections~\ref{sec:sem} and \ref{sec:soundnbe}).  The novel technical tool is a relation $T \shape S$ which relates a type $T$ to its possible shapes $S$.  This approximation relation allows reflection and reification at size-polymorphic types $\forall i.\, T$.
As usual for the meta-theory of Type Theory with large eliminations, the machinery is involved, but we just require the usual two logical relations:  First, a PER model to define the semantics of types and prove the completeness of NbE (Section~\ref{sec:sem}).  Secondly, a relation between syntax and semantics to prove soundness of NbE (Section~\ref{sec:soundnbe}).

\item We present an bidirectional type checking algorithm \cite{coquand:type} which takes the irrelevant size argument as reliable hint for the type checker (sections~\ref{sec:asub} and \ref{sec:tycheck}).  It is complete for normal forms which can be typed with the restricted rule for $\forall$-elimination:
\[
\ru{\Gamma \der t : \forall i.\, T
  }{\Gamma \der t\,a : T[a/i]}
\]
The algorithm employs the usual lazy reduction for types, \ie, just-in-time weak-head evaluation, in type and sub-type checker
\cite{huet:wsscs89}.  In this, it improves on \citet{fridlenderPagano:tlca13} which instruments full normalization (NbE) at every step.
\end{enumerate}
This article is accompanied by a prototypical type checker \textsf{Sit} which implements the type system and type checking algorithm as described in the remainder of the paper.
But before going into the technical details, we will motivate our type system from a practical perspective: reasoning about programs involving sized types in Agda.
